What to simulate? Inferring the direction for mental rotation

### 2014-01-31 man/cogsci-2014/cogsci-2014.tex
Finish first whole draft!

\usepackage{cogsci}
\usepackage{pslatex}
\usepackage{apacite}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{subfig}

@@

\input{../../results/D/overall_response_time.tex}
\input{../../results/D/overall_accuracy.tex}
\input{../../results/D/num_chance.tex}
\input{../../results/D/trial_accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}

@@

\begin{abstract}
\TODO{}

@@

\TODO{say something here}

\subsection{Mechanisms behind mental rotation}

@@

The only model (of which the authors are aware) that has seriously
attempted to address the decision of \textit{what} to simulate is the
parallel-process model of mental rotation by \citeA{Funt:1983wn}. He
determines the direction, axis, and extent of rotation by computing
the objects' principal axes using moments of inertia. One object is
then rotated through the target rotation, and is checked against the
other object for congruency. However, this approach assumes that the
corresponding points on the two objects can be easily identified,
which is not necessarily the case.

Perhaps people do not engage in a single trajectory of rotation that
ends with congruency. Some evidence suggests that people do not
perform a holistic rotation until the objects are congruent, but
rather pay attention to certain salient features of the object
\cite{Just1976,Yuille:1982tx}. A more recent study shows that when
performing \textit{physical} rotations, people do not rotate until
congruency is reached, and may even rotate \textit{away} from near
perfect matches \cite{Gardony:2013gn}. Moreover, the state-of-the-art
in computer vision suggests that there is more to shape matching than
simply checking for congruency, particularly when shapes are complex
or when the shapes might not be exactly the same
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}.

@@

If people are not performing single-trajectory rotations, what might
they be doing? To answer this question, we perform a ``rational
analysis'' of the problem of mental rotation
\cite{Marr:1983to,anderson90,Shepard:1987tt}. At the computational
level, we could say that the \textit{problem} is to determine
something about the spatial transformations that an object has
undergone. At the algorithmic level, we are constrained by the notion
that mental images must be transformed in an analog manner (or in a
way that is at least approximately analog), and that mental images are
time-consuming and effortful to generate. Thus, the \textit{goal} is
to make this determination while performing a minimum amount of
computation (i.e., as few rotations as possible).

The original congruency hypothesis \cite{Shepard1971} actually is a
rational solution which achieves this goal, in the sense that the
smallest amount of computation coincides with rotating through the
minimum angle. However, it relies on assumptions of \textit{a priori}
knowledge (either of the minimum angle or the correspondence of points
between shapes). Noting that a rational solution need not necessarily
maintain a single trajectory of rotation, we explore an alternative
rational model, which--rather than assuming the angle of rotation
\textit{a priori}--engages in an ``active sampling'' strategy.

@@

In the mental rotation task, people are presented with pairs of
two-dimensional shapes similar to those used by \citeA{Cooper:1975wp};
example stimuli are shown in Fig.~\ref{fig:model-traces}. The goal is
to determine whether the two images depict the same shape, or
mirror-image (flipped) shapes.

@@

We do not know which vertices of $\Xb$ correspond to which vertices of
$\Xa$, so $S$ must marginalize over the set of possible mappings,
$\mathbb{M}$. For brevity, let $\mathbf{X}_m=\M\cdot{}f(\Xa, h,
\theta)$ where $\M\in\mathbb{M}$ is a permutation matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{|\mathbb{M}|} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma_S^2),
  \label{eq:similarity}
\end{equation}
where $\sigma_S=0.15$ is the standard deviation of the similarity.

@@

One hypothesis in the literature is that people perform a mental
rotation in the shortest direction until a match is found
\cite{Shepard1971, Cooper:1975wp}, and that determining this direction
is an implementation detail \cite{Anderson1978}, or that it is
computed beforehand \cite{Funt:1983wn}.  To reflect this hypothesis,
we created an ``oracle'' model which first computes the correct
rotation and then simply takes actions in the shortest direction until
it reaches the point where the shapes are aligned.

@@

Formally, we denote $Z_h$ as our estimate of the likelihood for
hypothesis $h$, and write its distribution as:
\begin{equation}
  p(Z_h) = \int \left[\int S(\Xb, f(\Xa, \theta, h))p(\theta)\dif\theta\right] p(S)\dif S.
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use a Gaussian Process (GP) prior for $p(S)$, as in
\citeA{Osborne:2012tm}.\footnote{Because our domain is on a circle, we
  utilize a periodic kernel for the covariance
  \cite{Rasmussen:2006vz}, i.e. $k(\theta,
  \theta^\prime)=h^2\exp[-\frac{2}{w^2}\sin^2(\frac{1}{2}(\theta-\theta^\prime))]$.}
Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$, we first place a GP prior on
the log of $S_h$ in order to enforce positivity after it is
exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP, which we denote as
$\bar{S}_h:=\exp(\mu_h)$.\footnote{Note that the approach used here
  differs slightly from the one presented in \citeA{Osborne:2012tm}
  and is based on personal communication with one of the authors
  \cite{Duvenaud:2013td}.} Then, from \citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

@@

\begin{figure}[t]
  \centering
  \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_accuracy.pdf}}%
  \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_time.pdf}}
  \caption{\textbf{Effect of learning.} There was a clear effect of
    learning on participants' responses, especially in the first half
    of the experiment. Left: accuracy as a function of trial. Right:
    response time in milliseconds as a function of trial.}
  \label{fig:learning}
\end{figure}

@@

\subsection{Stimuli}

@@

\subsection{Participants and Design}

@@

All participants saw the same 10 practice trials, as described
above. There were 720 unique experimental stimuli (20 shapes $\times$
18 angles $\times$ 2 same/flipped), though because stimuli with
rotations of $0^\prime$ or $180^\prime$ were repeated twice, there
were 800 total experimental stimuli. These stimuli were split across 8
conditions in the following manner: first, stimuli were split into
four blocks of 200 trials. Within each block, each shape was repeated
10 times and each rotation was repeated 10 times (5 same, 5 flipped),
such that across all blocks, each stimulus appeared exactly once. Each
block was then split in half. Participants completed two half-blocks
(not necessarily both from the original full block).

\subsection{Procedure}

@@

On each trial, participants were instructed to press the `b' key to
begin and to focus on the fixation cross that appeared for 750
milliseconds afterwards. The two images were then presented
side-by-side, each at 300px $\times$ 300px, and participants could
press `s' to indicate they thought the images depicted the same
object, or `d' to indicate they thought the images depicted flipped
objects.

@@

\subsection{Analysis}

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a previous version
of the experiment, and 31 (12.6\%) because they failed a comprehension
check. A participant was marked as passing the comprehension check if
they answered at least 85\% of ``easy'' trials correctly. ``Easy''
trials were defined to be those for which the presented shapes
differed by a rotation of no more than 20 degrees. Additionally, we
excluded 82 individual trials (out of 40000) for which the response
time was either less than 100 milliseconds or greater than 20 seconds.

\subsection{Practice effects}

There was a significant effect of learning both on response time
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second
half). Fig.~\ref{fig:learning} shows scatterplots illustrating these
trends.

\subsection{Response times}

% The average response time across all stimuli was \ExpTime{} seconds.
% As expected, the minimum angle of rotation was significantly
% monotonically correlated with average per-stimulus response times,
% both for flipped pairs (\ExpThetaTimeCorrFlipped{}) and same pairs
% (\ExpThetaTimeCorrSame{}). This was also the case for the number of
% actions taken by the \Oc{} model (\OcThetaTimeCorr{}), the \Th{} model
% (\ThThetaTimeCorr{}), and the \Bq{} model (\BqThetaTimeCorr{}), but
% not the \Hc{} model (\HcThetaTimeCorr{}). The top row of
% Fig.~\ref{fig:response-time-accuracy} shows the overall effect of
% rotation on response times.

% The number of actions taken by the \Oc{} model was the best predictor
% of human response times, with a correlation of \ExpOcTimeCorr{}. The
% \Th{} model \Bq{} model with $p(h=0)=0.55$ were the next best (for
% \Th{}, \ExpThTimeCorr{}; for \Bq{}, \ExpBqpTimeCorr{}). Not far behind
% was the BQ model with equal prior (\ExpBqTimeCorr{}). The \Hc{} model
% was barely correlated with people (\ExpHcTimeCorr{}).

% Across all stimuli, participants' response times for ``flipped'' pairs
% were significantly higher than those for ``same'' pairs
% \TODO{confidence intervals} \TODO{flipped response times higher than
%   same response times for smaller angles?}

\subsection{Accuracy}

% The average accuracy across all stimuli was \ExpAccuracy{}, though
% there were \ExpNumChance{} individual stimuli (out of 720) for which
% people were not above chance. Of the models, the \Oc{} and \Th{}
% models performed perfectly, with 100\% accuracy. The \Bq{} model was
% very accurate overall (equal prior: \BqAccuracy{}, unequal prior:
% \BqpAccuracy{}). With the equal prior, there was \BqNumChance{}
% stimulus for which it was not above chance; with the unequal prior,
% there were \BqpNumChance{}. The \Hc{} model was not particularly
% accurate, and although it did perform above chance overall
% (\HcAccuracy{}), there were \HcNumChance{} individual stimuli for
% which it was not above chance.

% The minimum angle was also correlated with participants' average
% per-stimulus accuracy: for flipped pairs,
% \ExpThetaAccuracyCorrFlipped{}, and for same pairs,
% \ExpThetaAccuracyCorrSame{}. There was also a significant effect for
% the \Hc{} model (\HcThetaAccuracyCorr{}) and the \Bq{} model (equal
% prior: \BqThetaAccuracyCorr{}, unequal prior:
% \BqpThetaAccuracyCorr{}). The bottom row of
% Fig.~\ref{fig:response-time-accuracy} shows the effect of angle on
% accuracy for both participants and models.

% There was a moderate correlation between the accuracy of the \Hc{}
% model and human accuracy (\ExpHcAccuracyCorr{}). The \Bq{} model had a
% similar correlation (equal prior: \ExpBqAccuracyCorr{}, unequal prior:
% \ExpBqpAccuracyCorr{}). Correlations for the other models were
% undefined, because they had 100\% accuracy.

\begin{figure*}[t]
  \begin{center}
    \subfloat{\includegraphics[width=0.33\textwidth]{../../figures/D/theta_time_corrs.pdf}}
    \subfloat{\includegraphics[width=0.33\textwidth]{../../figures/D/theta_accuracy_corrs.pdf}}
    \subfloat{\includegraphics[width=0.33\textwidth]{../../figures/D/response_time_corrs.pdf}}
    \caption{}
    \label{fig:response-time-corrs}
  \end{center}
\end{figure*}

@@

We replicated the previous monotonic response time result, though not
the linear result. Actually kind of similar to \cite{Gardony:2013gn}
\TODO{elaborate}

Also replicated the constant accuracy for flipped stimuli, but not for
same stimuli, from \cite{Cooper:1975wp}. Why is this? Maybe the
strategy is ``they are different until proven the same''.

None of the models were particularly good fits to human data. The best
fit was the \Oc{} model, though this is not satisfying because the
\Oc{} model knows the answer beforehand, which people obviously do
not. The next best model, the \Th{} model, is also not a
particularly satisfying account, because it knows the global maximum
\textit{a priori}, which is not a reasonable assumption. \TODO{explain
  why it is faster for 180 than mid angles -- it is because for a lot
  of stimuli, there is rotational symmetry at 180, so it almost always
  starts checking the correct version first. For mid angles, it might
  check the wrong version first, meaning it takes much longer.}

The \Bq{} model makes more reasonable assumptions about the observer's
knowledge, but has its own problems. Because it is forced to take
small steps, it often gets stuck sampling the same region over and
over again, particularly for stimuli with larger rotations
(Fig.~\ref{fig:model-traces}). As a result, its response time curve is
convex rather than concave.

Other problems: evidence that rotations aren't actually holistic
\cite{Yuille:1982tx}, eye movements \cite{Just1976}

\TODO{talk about how these results suggest a model of mental rotation
  is harder than was previous thought, and that this is a first step
  towards detangling these issues.}

>>>>

\usepackage{cogsci}
\usepackage{pslatex}
\usepackage{apacite}
\usepackage{graphicx}
\usepackage{color}
\usepackage{amsmath, amsthm, amssymb}
\usepackage{subcaption}
\usepackage{caption}

@@

\input{../../results/D/overall_response_time.tex}
\input{../../results/D/overall_accuracy.tex}
\input{../../results/D/num_chance.tex}
\input{../../results/D/trial_accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}
\input{../../results/D/response_time_corrs.tex}
\input{../../results/D/accuracy_corrs.tex}

@@

\begin{abstract}
  When people use mental imagery to solve tasks, how do they decide
  \textit{which} images to generate? To answer this question, we
  explored how mental simulation should used in the classic
  psychological task of determining if two images depict the same
  object in different orientations \citeA{Shepard1971}. Through a
  rational analysis of mental rotation, we formalized four models and
  compared them to human performance on a 2D mental rotation
  task. None of the models were a particularly satisfying explanation
  of human behavior, leading us to suggest that the question of ``what
  to simulate?'' is more difficult than has previously been
  assumed. While only exploratory, this work provides the foundation
  necessary to begin answering this question.

@@

@@

The only models (of which the authors are aware) that have seriously
attempted to address the decision of \textit{what} to simulate are
those by \citeA{Funt:1983wn} and \citeA{Just:1985uu}. In both of these
models, the axis and direction of rotation are computed prior to
performing the rotation. One object is then rotated through the target
rotation, and is checked against the other object for
congruency. However, this approach assumes that the corresponding
points on the two objects can be easily identified, which is not
necessarily the case. Indeed, the state-of-the-art in computer vision
suggests that there is more to shape matching than simply checking for
congruency, particularly when shapes are complex or when the shapes
might not be exactly the same
\cite<e.g.,>{Belongie:2002tj,Sebastian:2003vm}. Additionally, recent
research shows that when performing \textit{physical} rotations,
people do not rotate until congruency is reached, and may even rotate
\textit{away} from near perfect matches \cite{Gardony:2013gn}.

@@

If people are not computing the rotation beforehand, what might they
be doing? To answer this question, we perform a ``rational analysis''
of the problem of mental rotation
\cite{Marr:1983to,anderson90,Shepard:1987tt}. At the computational
level, we could say that the \textit{problem} is to determine which
spatial transformations an object has undergone based on two
depictions of that object which do not include information about point
correspondences across images. At the algorithmic level, we are
constrained by the notion that mental images must be transformed in an
analog manner (or in a way that is at least approximately analog), and
that mental images are time-consuming and effortful to generate. Thus,
the \textit{goal} is to make this determination while performing a
minimum amount of computation (i.e., as few rotations as possible).

The original congruency hypothesis \cite{Shepard1971} actually is a
rational solution which achieves this goal, in the sense that the
smallest amount of computation coincides with rotating through the
minimum angle. However, it violates the constraint that we do not know
points correspondences, which is exactly what necessitates the use of
imagery. Noting that a rational solution need not necessarily maintain
a single trajectory of rotation, we explore an alternative rational
model, which--rather than computing the angle of rotation ahead of
time--engages in an ``active sampling'' strategy.

@@

In the mental rotation task, people are presented with pairs of
two-dimensional shapes similar to those used by \citeA{Cooper:1975wp};
example stimuli are shown in Fig.~\ref{fig:model-traces}. The goal is
to determine whether the two images depict the same shape, or
mirror-image shapes.

@@

We do not know which vertices of $\Xb$ correspond to which vertices of
$\Xa$, so $S$ must marginalize over the set of possible mappings. For
brevity, let $\mathbf{X}_m=\M\cdot{}f(\Xa, h, \theta)$ where $\M$ is a
permutation matrix. Then:
\begin{equation}
  S(\Xb, f(\Xa, h, \theta)):=\frac{1}{2N} \sum_{\M} \prod_{n=1}^N \mathcal{N}(\mathbf{x}_{bn}\ \vert \ \mathbf{x}_{mn}, \I\sigma_S^2),
  \label{eq:similarity}
\end{equation}
where $N$ is the number of vertices, $2N$ is the total number of
possible mappings,\footnote{It is $2N$ and not $N^2$ because, in polar
  coordinates, vertices are always connected to their two nearest
  neighbors in the $\theta$ dimension.}and $\sigma_S=0.15$ is the
standard deviation of the similarity.

@@

One hypothesis in the literature is that people perform a mental
rotation in the shortest direction until a match is found
\cite{Shepard1971, Cooper:1975wp}, and that determining this direction
is an implementation detail \cite{Anderson1978}, or that it is
computed beforehand \cite{Funt:1983wn}.  To reflect this hypothesis,
we created an ``oracle'' model which is told which points on each
shape correspond. The \Oc{} model first computes the correct rotation
and then simply takes actions in the shortest direction until it
reaches the point where the shapes are aligned.

@@

Formally, we denote $Z_h$ as our estimate of the likelihood for
hypothesis $h$, and write its distribution as:
\begin{equation}
  p(Z_h) = \int \left[\int S(\Xb, f(\Xa, \theta, h))p(\theta)\dif\theta\right] p(S)\dif S.
\end{equation}
This is a technique known in the machine-learning literature as
\textit{Bayesian Quadrature} (BQ)
\cite{Diaconis:1988uo,OHagan:1991tx,Osborne:2012tm}. Specifically, we
use a Gaussian Process (GP) prior for $p(S)$, as in
\citeA{Osborne:2012tm}.\footnote{Because our domain is on a circle, we
  utilize a periodic kernel for the covariance
  \cite{Rasmussen:2006vz}.}  Denoting $S_h=S(\Xb, f(\Xa, \theta, h))$,
we first place a GP prior on the log of $S_h$ in order to enforce
positivity after it is exponentiated:
\begin{equation}
\mathbb{E}[Z_h] \approx \int \exp(\mu_h(\theta))p(\theta)\dif\theta,
\end{equation}
where $\mu_h:=\mu(\log S_h)$ is the mean of the log-GP. However, this
integral is still intractable. To approximate it, we fit a second GP
over points sampled from the log-GP, which we denote as
$\bar{S}_h:=\exp(\mu_h)$.\footnote{Note that the approach used here
  differs slightly from the one presented in \citeA{Osborne:2012tm}
  and is based on personal communication with one of the authors
  \cite{Duvenaud:2013td}.} Then, from \citeA{Duvenaud:2013td}:
\begin{align}
  \mathbb{E}[Z_h] &\approx \int \bar{\mu}_h(\theta)p(\theta)\dif\theta,\\
  \mathbb{V}(Z_h) &\approx \iint \mathrm{Cov}_h(\theta,
  \theta^\prime)\bar{\mu}_h(\theta)\bar{\mu}_h(\theta^\prime)p(\theta)p(\theta^\prime)\dif\theta\dif\theta^\prime,
\end{align}
where $\bar{\mu}_h:=\mu(\bar{S}_h)$ is the mean of the second GP, and
$\mathrm{Cov}_h:=\mathrm{Cov}(\log S_h)$ is the covariance of the
log-GP.

@@

% \begin{figure}[t]
%   \centering
%   \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_accuracy.pdf}}%
%   \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_time.pdf}}
%   \caption{\textbf{Effect of learning.} There was a clear effect of
%     learning on participants' responses, especially in the first half
%     of the experiment. Left: accuracy as a function of trial. Right:
%     response time in milliseconds as a function of trial.}
%   \label{fig:learning}
% \end{figure}

@@

\subsubsection{Stimuli}

@@

\subsubsection{Participants and Design}

@@

All participants saw the same 10 practice trials, as described
above. There were 720 unique experimental stimuli (20 shapes $\times$
18 angles $\times$ 2 reflections), though because stimuli with
rotations of $0^\prime$ or $180^\prime$ were repeated twice, there
were 800 total experimental stimuli. These stimuli were split across 8
conditions in the following manner: first, stimuli were split into
four blocks of 200 trials. Within each block, each shape was repeated
10 times and each rotation was repeated 10 times (5 ``same'', 5
``flipped''), such that across all blocks, each stimulus appeared
exactly once. Each block was then split in half. Participants
completed two half-blocks (not necessarily both from the original full
block).

\subsubsection{Procedure}

\begin{figure*}[t]
  \centering
  \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[width=\textwidth]{../../figures/D/theta_time_corrs.pdf}
    \caption{\textbf{Angle vs. response time}}
    \label{fig:theta-time-corrs}
  \end{subfigure}%
  \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[width=\textwidth]{../../figures/D/theta_accuracy_corrs.pdf}
    \caption{\textbf{Angle vs. accuracy}}
    \label{fig:theta-accuracy-corrs}
  \end{subfigure}%
  \begin{subfigure}[b]{0.32\textwidth}
    \includegraphics[width=\textwidth]{../../figures/D/human_model_corrs.pdf}
    \caption{\textbf{Human vs. model}}
    \label{fig:human-model-corrs}
  \end{subfigure}
  \vspace{10pt}
  \caption{\textbf{Correlations.} (a) Spearman correlations between
    the minimum angle of rotation and response time. (b) Spearman
    correlations between the minimum angle of rotation and
    accuracy. The \Oc{} and \Th{} models had 100\% accuracy;
    therefore, their correlations are undefined. (c) Pearson
    correlations between human and model, both for response time and
    accuracy. All error bars are 95\% confidence intervals.}
\end{figure*}

@@

On each trial, participants were instructed to press the `b' key to
begin and to focus on the fixation cross that appeared for 750
milliseconds afterwards. The two images were then presented
side-by-side, each at 300px $\times$ 300px, and participants could
press `s' to indicate they thought the images depicted the ``same''
object, or `d' to indicate they thought the images depicted
``flipped'' objects.

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a previous version
of the experiment, and 31 (12.6\%) because they failed a comprehension
check. A participant was marked as passing the comprehension check if
they answered at least 85\% of ``easy'' trials correctly. ``Easy''
trials were defined to be those for which the presented shapes
differed by a rotation of no more than 20 degrees. Additionally, we
excluded 82 trials for which the response time was either less than
100 milliseconds or greater than 20 seconds.

\subsubsection{General analysis}

@@

\subsubsection{Human}

The average response time across all stimuli was \ExpTime{} seconds.
As illustrated in Fig.~\ref{fig:theta-time-corrs}, the minimum angle
of rotation was significantly monotonically correlated with average
per-stimulus response times, both for ``flipped'' and ``same''
pairs. While this replicates the general result of previous
experiments \cite<e.g.,>{Shepard1971,Cooper:1975wp}, our results are
not as perfectly linear. However, \citeA{Gardony:2013gn} found a
constant effect for ``flipped'' pairs; an average of their results and
the classic results would produce something similar to ours.

The average accuracy across all stimuli was \ExpAccuracy{}, though
there were \ExpNumChance{} individual stimuli (out of 720) for which
people were not above chance.  The minimum angle was also correlated
with participants' average per-stimulus accuracy, as shown in
Fig.~\ref{fig:theta-accuracy-corrs}, though significantly more so for
``same'' pairs. This is the same result found both by
\citeA{Cooper:1975wp} and \citeA{Gardony:2013gn}.

There was a significant effect of learning both on response time
(\ExpTrialTimeCorr{}) and on accuracy (\ExpTrialAccuracyCorr{}),
though the effect on accuracy was not significant during the second
half of the experiment (\ExpaTrialAccuracyCorr{} for the first half
vs. \ExpbTrialAccuracyCorr{} for the second half). These practice
effects may have contributed to the not-quite-linearity of the human
response times; future work will need to collect more data per
participant to mitigate this effect.

\subsubsection{\Oc{} model}

As we would expect, the number of actions taken by the \Oc{} model was
perfectly correlated with the minimum angle of rotation
(Fig.~\ref{fig:theta-time-corrs}), both for ``same'' and ``flipped''
pairs. Unsurprisingly, then, \Oc{} model was the best fit to human
response times, with a correlation of \OcTimeCorr{}
(Fig.~\ref{fig:human-model-corrs}).  However, the \Oc{} model was
100\% accurate, and therefore the correlation between minimum angle
and accuracy was undefined (Fig.~\ref{fig:theta-accuracy-corrs}).

\subsubsection{\Th{} model}

The number of actions taken by the \Th{} model was about as correlated
with the minimum angle as human response times were
(Fig.~\ref{fig:theta-time-corrs}), though this effect was less strong
for individual shapes (e.g.,
Fig.~\ref{fig:response-time-stimulus}). Like the \Oc{} model, the
\Th{} model had 100\% accuracy, and thus there was no correlation
between minimum angle and accuracy. The \Th{} was able to explain a
moderate amount of the variance in human response times, with a
correlation of \ThTimeCorr{} (Fig.~\ref{fig:human-model-corrs}).

\subsubsection{\Hc{} model}

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{../../figures/D/response_time_stimulus.pdf}
    \caption{\textbf{Typical response time curves for a single shape.}
      These plots correspond to the shape shown in
      Fig.~\ref{fig:model-traces}. Left: human curves are either
      linear (as with the ``same'' pairs), or linear and then flat (as
      with the ``flipped'' pairs). Middle: the \Th{} model tends to
      have highly inconsitent within-shape response times. Right: the
      \Bq{} model is roughly linear.}
    \label{fig:response-time-stimulus}
  \end{center}
\end{figure}

The \Hc{} was the only model for which there was no effect of rotation
on response time (Fig.~\ref{fig:theta-time-corrs}). Although there was
a significant effect of rotation on accuracy
(Fig.~\ref{fig:theta-accuracy-corrs}), the \Hc{} model was barely
above chance (\HcAccuracy{}) and there were \HcNumChance{} individual
stimuli for which it was not above chance. The \Hc{} model was not a
particularly good predictor of human response times (\HcTimeCorr{}),
though it was a moderate predictor of accuracy (\HcAccuracyCorr{}), as
shown in Fig.~\ref{fig:human-model-corrs}.

\subsubsection{\Bq{} model}

As with most of the other models, there was a significant effect of
rotation on the number of steps taken by the \Bq{} model
(Fig.~\ref{fig:theta-time-corrs}). The \Bq{} model explained variance
in human response times about as well as the \Th{} model
(Fig.~\ref{fig:human-model-corrs}), with a correlation of
\BqTimeCorr{} for the equal prior and \BqpTimeCorr{} for the unequal
prior.

The \Bq{} model was quite accurate overall, both with the equal prior
(\BqAccuracy{}) and unequal prior (\BqpAccuracy{}). With the equal
prior, there was \BqNumChance{} stimulus for which it was not above
chance; with the unequal prior, there were \BqpNumChance{}. There was
a significant effect of rotation on the \Bq{} model's accuracy
(Fig.~\ref{fig:theta-accuracy-corrs}), and itt explained about the
same amount of variance in people's accuracy as the \Hc{} model: for
the equal prior, \BqAccuracyCorr{}; for the unequal prior,
\BqpAccuracyCorr{}.

Because the \Bq{} model relied on Eq.~\ref{eq:lh-ratio} for its
stopping criteria (as opposed to just finding a maximum), the prior
$p(h)$ had a fairly significant effect, as shown in
Fig.~\ref{fig:response-time-accuracy}. With just a small bias of
$p(h=0)=0.55$, there is a clear separation in response times that is
similar to the separation seen in human response times. The prior also
had an effect on accuracy, though this effect did not reflect the
trends in human behavior.

@@

We set out to answer the question: how do people decide \textit{what}
to simulate when using mental imagery? Focusing on the specific case
of determining the direction and extent of mental rotation, we
formalized four models and compared their performance with the results
of a behavioral experiment. The \Oc{} model was the best fit to human
response times, but failed to capture anything regarding accuracy. The
\Th{} model was a slightly worse fit to human response times, and also
did not capture any trends in accuracy. The \Bq{} model fit human
reponse time data as well as the \Th{} model, and captured a small
amount of the variance in people's accuracy. The \Hc{} model barely
fit human response time data, though it did fit accuracy as well as
the \Bq{} model with an equal prior.

While it is clear that none of the models fully capture human behavior
on the mental rotation task, it is worth noting where they shed light
on aspects that have previously gone unexplained.  For example, the
\Bq{} model's adaptive stopping rule is sensitive to the prior, and
thus provides a possible explanation for why people are slower to
respond on ``flipped'' stimulus pairs.

Additionally, a closer look at individual shapes reveals several
interesting points (Fig.~\ref{fig:response-time-stimulus} shows curves
for one particular shape; these curves are fairly typical relative to
other shapes). First, the curves for the human response times on
specific shapes tend to either be linear, or linear first, then
flat. Second, while the \Th{} model appears to capture the concavity
of human response times overall, it is actually fairly inconsistent
within shapes. Third, the linearity of the \Bq{} model's response time
curves \textit{is} consistent within shapes, though it does not
exhibit the same ``plateau'' effect that sometimes appears in the
human curves.

How could the models we evaluated be improved? First, all of the
models use holistic rotations, however there is evidence that people
might compare individual features of shapes rather than the entire
shape at once \cite{Just1976,Yuille:1982tx}. Second, there are other
possible ``active sampling'' approaches that could combine the best
features from the \Th{} and \Bq{} models, such as maintaining a
distribution over the location and value of the global maximum, rather
than over the integral. Third, studying rotations with 3D objects of
varying levels of complexity may be more informative than the
extremely simple 2D objects used in this experiment.

As a parting comment, we note that the question of how people decide
\textit{which} rotations to perform is more difficult to answer than
previously assumed. Although we examined four well-motivated models,
none came close to explaining the full range of human behavior on this
task. We suggest that such metacognitive questions are fertile areas
for future research.
